---
title: "Notes - Linear Models"
author: "Madelynn VanderHoek"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: TRUE
  md_document: 
---

# Notes on Linear Models: 03/24/2026

## Packages loaded
```{r}
library(tidyverse)
library(lme4)
library(emmeans)
library(multcomp)
```

## Introduction to Regression Analysis in R

In these notes, we will learn the basics of R’s capabilities in statistical tests from the perspective of the linear model. If you understand the linear model, you should understand other statistical tests since all (ANOVA, ANCOVA, Multivariable linear model, t-tests, generalized linear models, mixed effect linear models, correlation), are unified under the concepts of a linear model. 


If you needed more information on any material covered here and much more precise explanations than I could ever provide, please see [Dr. Steury's lectures here](https://webhome.auburn.edu/~tds0009/wild7150.html)

For these Notes, we are concerned with the *basic linear equation*.

In a standard math class you may have learned in high-school, we would write this as: y = mx + b
-   Where y is our response, m is our slope, x is the other variable, and b is the intercept.

In statistics we write this as: y(hat) = β0 + β1x
-   where y(hat) = equation - β0 = intercept - β1 = slope - x = x points

The actual modeled line is equal to the equation, which adds on the error term. yi = β0 + β1xi + ε~N(0,σ)
-   where yi (y sub i) = each point on the y axis - β0 = intercept - β1 = slope - xi = each x point - then ε = error - and N(0,σ) = normal distribution with a mean 0 and standard deviation = sigma


-   The goal of all of this is to estimate three terms. - slope - intercept - standard deviation
-   The goal is to minimize the distance from each point to the line. In statistical terms, we want to minimize the sum of squared errors or SSE. The sum of   squares of the regression (SSR) is the distance between the best fit line and the average. Adding these together equals the total sum of squares. Partitioning this into SSR and SSE is related to pvalue. The smaller the SSE and higher the SSR, the smaller the p-value.
-   Then, we want to know if it is statistically significant. What is the chance we get the result? - that is your p-value.
-   The null hypothesis is that H0 = No slope or that β1 = 0
-   In general: - as slope goes up p-value goes down - as sample size goes up p-value goes down - as noise (error) goes down p-value goes down


## Continuous X and Continuous Y 

Load the mtcars data set that is in bedded in R to help us demonstrate basic linear regression:

The variables inside the cars data set include:
-   mpg = the mpg
-   cyl = number of cylinders
-   disp = displacement
-   hp = gross horsepower
-   drat = rear axle ratio
-   wt = weight
-   qsec = quarter mile time
-   vs = engine shape V = 0 or Straight = 1
-   am = Transmission type; 0 = auto, 1 = manual
-   gear = number of forward gears
-   carb = number of carburetors

```{r}
data("mtcars")
```

The first step is to always *visualize your data*

-   If you have two continuous variables, you will want to visualize via a scatter plot. 
```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_smooth(method = lm, se = FALSE, color = "grey") + #geom_smooth() using formula = 'y ~ x'
  geom_point(aes(color = wt)) +
  xlab("Weight") + 
  ylab("Miles per gallon") +
  scale_colour_gradient(low = "forestgreen", high = "black") +
  theme_classic()
```

Does it look like there is a relationship? Absolutely! But how do we know?
 
*Next Step:* We run a linear model - or if it's a continuous x variable and a continuous y variable, we would call it a regression. If we consider it a cause-and-effect relationship, we may call it a correlation.
-   The simplest way to do this is with the function lm()


```{r}
lm(mpg~wt, data = mtcars)
```

*Results:* When we run this it gives us an output of the predicted Intercept or our β0 parameter.

*Next Step:* We can now run summary of this linear model to output some summary statistics

```{r}
summary(lm(mpg~wt, data = mtcars))
```

*Results:* Gave us apretty good idea that the slope is not equal to 0, and the intercept estimate is pretty good.
-   Another thing to pay attention to is the R squared. This tells you the variation in y explained by x. So, in our case, about 74% of the variation in y is explained by the x variable.

*Next Step:* Ok, what if we wanted an ANOVA table from this? We could run an ANOVA.

```{r}
anova(lm(mpg~wt, data = mtcars))
```

*Results:* Hey, the p-value is the same!!! It means that a linear model and an ANOVA are essentially the same things, and the value reported in the ANOVA is the value of the linear regression or our slope parameter. This means that wt has a significant effect on miles per gallon.

*Next Steps:* We may want to view this in a correlation.

```{r}
cor.test(mtcars$wt, mtcars$mpg)
```

*Results:* The p-value is the same again! This gives you a different r value, which is the correlation statistic. The closer to -1 or 1, the stronger the correlation between the two variables.

## Assumptions

In general, there are several assumptions in a regression, linear model, ANOVA, or whatever you want to call it.

They are:

-   y is continuous
-   error is normally distributed 
-   relationship is linear 
-   homoskedasticity 
-   sigma is consistent 
-   independent samples

However, regression/anovas/linear models are generally robust enough to moderate departures from the assumptions. So, do the assumptions matter… yeah, but only if they are terrible. However, in most cases, if one assumption is violated, it won't change the result too much. If anything, it will increase the p-value, and your conclusion is more conservative at that point.

Yes, there are assumption tests, but we will not do them. You just need to know how to read your data and if you want look at the residual plot to diagnose your violated assumptions.

In our example above we can simply get the residuals from our linear model like this

```{r}
model <- lm(mpg~wt, data = mtcars)

ggplot(model, aes(y = .resid, x = .fitted)) +
  geom_point() +
  geom_hline(yintercept = 0)
```


## Categorical Variables 

Now if we have a categorical x variable and a continuous y variable we would perform a *t-test*.

Load the fungicide study bull richness data set from previous notes for this example: 
```{r}
bull.rich <- read.csv("Bull_richness.csv")
```

Filter our data set to include one treatment and growth stage for demonstration of a t-test.

```{r}
bull.rich %>%
  filter(GrowthStage == "V8" & Treatment == "Conv.") %>%
  ggplot(aes(x = Fungicide, y = richness)) + 
  geom_boxplot()
```

*Results:* Does it look like there is a difference, yeah! But how do we know?

*Next Steps:* We can do a t-test. Where the null hypothesis is that there is no difference in group means

```{r}
bull.rich.sub <- bull.rich %>%
  filter(GrowthStage == "V8" & Treatment == "Conv.")

t.test(richness~Fungicide, data = bull.rich.sub)
```

*Results:* What is actually happening mathematically here?
-   Remember our equation? yi = β0 + β1xi + ε~N(0,σ)
-   If we code Control (C) as a 0 and fungicide (F) as 1 and plug it into the equation as x we end up with: richness = β0 + 0 + ε~N(0,σ) and richness = β0 + β1 + ε~N(0,σ)
-   β0 is still the intercept but it turns into the distance between the mean of the Control group and 0 and the slope β1 is equal to the difference between groups.
-   Visually this is what we are looking at. The t-test is a regression!

*Next Steps:* We can plug this into our linear model and get the same results.
```{r}
t.test(richness~Fungicide, data = bull.rich.sub, var.equal = TRUE)

summary(lm(richness~Fungicide, data = bull.rich.sub))

anova(lm(richness~Fungicide, data = bull.rich.sub))
```

Assuming equal variance in groups and performing a two-sample t-test is the same result as a linear model and ANOVA on the linear model. The p-value tells us that the slope of the line is different than 0.


## ANOVAs
ANOVAs would have continuous y and multinational categorical x. In other words, more than two groups.

Lets filter our data set to have only richness in different crop growth stages in the control samples in soybean in conventional management. Here we have three growth stages.
```{r}
bull.rich.sub2 <- bull.rich %>%
  filter(Fungicide == "C" & Treatment == "Conv." & Crop == "Corn")
```

Visualize the data
```{r}
bull.rich.sub2$GrowthStage <- factor(bull.rich.sub2$GrowthStage, levels = c("V6", "V8", "V15"))

ggplot(bull.rich.sub2, aes(x = GrowthStage, y = richness)) +
  geom_boxplot()
```


*Results:* Well, there looks to be a difference between the groups.
-   Mathematically, we again can think of this as a regression analysis where we have β0 as the intercept, or the difference between the mean of V6 and 0, V8 as β1, and V15 as β2.
-   Full equation: y = β0 + β1(V8) + β1(V15) + ε~N(0,σ)
-   Coding each term as 1 or 0 we get the following equations.
-   richness(V6) = β0 + 0 + ε~N(0,σ) and richness(V8) = β0 + β1 + ε~N(0,σ) richness(V15) = β0 + β2 + ε~N(0,σ)
-   Visually it looks like this, and guess what, it's a regression!

To show you lets do the actual test.
Set up the linear model and run various ANOVAs… holy smokes they all have the same information. Thats because an ANOVA is a regression!

```{r}
lm.growth <- lm(richness ~ GrowthStage, data = bull.rich.sub2)
summary(lm.growth)

anova(lm.growth)

summary(aov(richness ~ GrowthStage, data = bull.rich.sub2))
```

But how do we know which is different from the the other group?

We can do post-hoc tests to find out. The simplest way to think of this is individual t-tests across groups. The most versatile way to do this is with the packages means, and multcomp.

*Next Steps:* The lsmeans are the least squared means - the means estimated by the linear model. This contrasts the arithmetic means, which are the means calculated or the average.


```{r}
lsmeans <- emmeans(lm.growth, ~GrowthStage) # estimate lsmeans of variety within siteXyear
Results_lsmeans <- cld(lsmeans, alpha = 0.05, reversed = TRUE, details = TRUE) # contrast with Tukey ajustment by default. 
Results_lsmeans
```

## Interactions Terms

-   So far we have done one variable, but what if we care about the interactions between factors. In other words what if we care about if the effect of one factor depends on the other.
-   In our example, maybe we care about if the effect of fungicide depends on time.
-   We can do this within a linear model as well using the * between factors.


Lets filter our dataset to include fungicide term.
```{r}
bull.rich.sub3 <- bull.rich %>%
  filter(Treatment == "Conv." & Crop == "Corn")

bull.rich.sub3$GrowthStage <- factor(bull.rich.sub3$GrowthStage, levels = c("V6", "V8", "V15"))
```

Now lets set up our linear model with fungicide interaction factor:
```{r}
# write it like this
lm.inter <- lm(richness ~ GrowthStage + Fungicide + GrowthStage:Fungicide, data = bull.rich.sub3)

# or like this
lm(richness ~ GrowthStage*Fungicide, data = bull.rich.sub3)
```


Overall if we look at the summary we can see that there are significant terms and the overall p value is significant. If we want to look at the ANOVA table we can do this.
```{r}
summary(lm.inter) # significant terms

anova(lm.inter) # The interaction term is signifant. 
```

*Results:*So the interaction term is significant. In general you want to look at the most complex interaction that is significant and look at that first. But this means that the effect of fungicide is dependent on the growthstage, or in other words at least one comparison of fungicide within growthstage or growthstage within fungicide is significant.

*Next Steps:* For this study we more care about the effect of fungicide over the levels of growthstage so we can plot that easily.
```{r}
bull.rich.sub3 %>%
  ggplot(aes(x = GrowthStage, y = richness, fill = Fungicide)) +
  geom_boxplot()
```

*Results:* So, by visualization, we can see that it looks like the V8 fungicide affected richness more than others, and by V15, they recovered.

*Next Steps:* But how do we know? In this case, we have to do a post-hoc test within the levels of growthstage.

```{r}
lsmeans <- emmeans(lm.inter, ~Fungicide|GrowthStage) # estimate lsmeans of variety within siteXyear
Results_lsmeans <- cld(lsmeans, alpha = 0.05, reversed = TRUE, details = TRUE) # contrast with Tukey ajustment
Results_lsmeans
```

## Mixed Effects Models 
In mixed-effects models, we have fixed and random effects terms. The random effects term is something that affects the variation in y. A fixed effect is something that affects the mean of y. There are NO rules that determine what is a fixed or random effect. You have to justify that in your experimental design. Another way to think about this would be factors that you do not really care about affecting the mean but still want to account for that variation in your model so you can generalize over that factor. A common way that random effects are included in experimental designs is called the blocking factor.

-   However, common fixed and random effects will apply to folks in this class.

Common fixed effects
-   Treatment
-   Species
-   gene

Common random effects (blocking factor)
-   Year
-   replicate
-   trial
-   Individuals
-   Fields

Let's see some random effects models in action.

Let's use our interaction model from above.
```{r}
lme0 <- lm(richness ~ GrowthStage*Fungicide, data = bull.rich.sub3)
```

*Next Steps:* How do we include a random effect in this? Well… first, we have to use a different package, and for this, we are going to use the package lme4.

```{r}
lme1 <- lmer(richness ~ GrowthStage*Fungicide + (1|Rep), data = bull.rich.sub3)
summary(lme1)
```

Here we are including the between replicate variation (spatial field variation) as an error term and estimate the sigma (standard deviation) due to the error associated with Rep. 

*Next Steps:* What does this do? Well, let's look at what it does to our Betas in our original model.

```{r}
summary(lme0)

summary(lme1)
```

For some terms, it actually makes the standard error of our betas go down, which means we have better estimates of our betas, and we are more confident that we are estimating them correctly in our linear models, which means we have more power to detect differences.


[Link to LinearModelsNotes on GitHub](https://github.com/MadelynnV/PLAP-Clas-2026/tree/main/ClassNotes/LinearModels_folder)



