---
title: "Data Wrangling Notes"
author: "Madelynn VanderHoek"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
  md_document:
---






##### Notes about data Wrangling in tidyverse packages

Data set used for these notes:

```{r}
library(tidyverse)

microbiome.fungi <- read.csv("Bull_richness.csv")
str(microbiome.fungi)
```

To visualize only select columns of your data you can accomplish this by the select function.

```{r}
microbiome.fungi2 <- select(microbiome.fungi, SampleID, Crop, Compartment:Fungicide, richness)
head(microbiome.fungi2)
```

The filter() function allows you to subset the data fairly easily.

```{r}
head(filter(microbiome.fungi2, Treatment == "Conv."))
```

You can also perform this function with other symblos other then a comma

```{r}
head(filter(microbiome.fungi2, Treatment == "Conv." & Fungicide == "C"))
head(filter(microbiome.fungi2, Sample == "A" | Sample == "B"))
```


##### Creating a new Column within tidyverse package

-   The mutate function allows you to create new columns within your data set

```{r}
head(mutate(microbiome.fungi2, logRich = log(richness)))
```

You can also create an new column which combines Crop and Treatment

```{r}
head(mutate(microbiome.fungi2, Crop_Treatment = paste(Crop, Treatment )))
```


##### Creating an pipe line - which combine multiple function together

We can combine all of the previous steps into one large string of functions

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  filter(Treatment == "Conv.") %>% #subsetting to only include the conventional treatments
  mutate(logRich = log(richness)) %>% #creating a new column of the log richness
  head() 
```

##### Summaries data for basic statistics

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  filter(Treatment == "Conv.") %>% #subsetting to only include the conventional treatments
  mutate(logRich = log(richness)) %>% #creating a new column of the log richness
  summarise(Mean.rich = mean(logRich))
```

We can also multiple summary statistics using this function

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  filter(Treatment == "Conv.") %>% #subsetting to only include the conventional treatments
  mutate(logRich = log(richness)) %>% #creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), 
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n))
```

It might be helpful to have the summary statistics by group. We can do this by using the group_by() function.

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  group_by(Treatment, Fungicide) %>% #subsetting to only include the conventional treatments
  mutate(logRich = log(richness)) %>% #creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), 
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n))


```

##### Connecting to Plotting in ggplot

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  group_by(Treatment, Fungicide) %>% #subsetting to only include the conventional treatments
  mutate(logRich = log(richness)) %>% #creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), 
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n)) %>%
  ggplot(aes(x = Fungicide, y = Mean.rich)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(x = Fungicide, ymin = Mean.rich-std.err, ymax = Mean.rich+std.err), width = 0.4) +
  theme_minimal() +
  xlab("") +
  ylab("Log Richness") +
  facet_wrap(~Treatment)
```

##### Joining group of function

-   matching up data by rows based on some criteria.
-   left_join(): keep all rows of X and add matching rows from Y. Any rows n Y that don't match X are excluded.
-   right_join(), reverse of left_join(), inner_join(): only keep rows that are common to both X and Y, remove everything else
-   full_join(): keep any columns that are in either X or Y

To practice these function we are going to split data into separate sets, including taxonomy and metadata

```{r}
# selecting just the richness and sample ID
richness <- microbiome.fungi %>%
  select(SampleID, richness)

# selecting columns that dont include the richness
metadata <- microbiome.fungi %>%
  select(SampleID, Fungicide, Crop, Compartment, GrowthStage, Treatment, Rep, Sample)

head(richness)
head(metadata)

```

Now we are going to add the richness data to the metadata based on the common column of SampleID

```{r}
head(left_join(metadata, richness, by = "SampleID"))
```

##### Pivoting: changing the format of the data

-   pivot_longer()
-   pivot_wider()

Example of using pivot_wider() function

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>%
  group_by(Treatment, Fungicide) %>%
  summarise(Mean = mean(richness)) %>%
  pivot_wider(names_from = Fungicide, values_from = Mean)

```

With this function you can easily take the difference between the fungicide and control.

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>%
  group_by(Treatment, Fungicide) %>%
  summarise(Mean = mean(richness)) %>%
  pivot_wider(names_from = Fungicide, values_from = Mean) %>%
  mutate(diff.fungicide = C - F)


```

In a plot:

```{r}
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>%
  group_by(Treatment, Fungicide) %>%
  summarise(Mean = mean(richness)) %>%
  pivot_wider(names_from = Fungicide, values_from = Mean) %>%
  mutate(diff.fungicide = C - F) %>%
  ggplot(aes(x = Treatment, y = diff.fungicide)) +
  geom_col() +
  theme_minimal() +
  xlab("") +
  ylab("Difference n average species richness")

```
