---
title: "Coding Challenge 7"
author: "Madelynn VanderHoek"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
  github_document:  
---

# Coding Challenge 7 - 3/26/2026

## Question 1. 
Read in the data called “PlantEmergence.csv” using a relative file path and load the following libraries. tidyverse, lme4, emmeans, multcomp, and multcompView. Turn the Treatment, DaysAfterPlanting and Rep into factors using the function as.factor

```{r}
# example shown here: STAND$Treatment <- as.factor(STAND$Treatment) 
```

```{r}
# Read in the data:
PlantEmergence <- read.csv("PlantEmergence.csv")

#Load Librarys:
library(tidyverse)
library(lme4)
library(emmeans)
library(multcomp)
library(multcompView)

#Using the as.function()
PlantEmergence$Treatment <- as.factor(PlantEmergence$Treatment)
PlantEmergence$DaysAfterPlanting <- as.factor(PlantEmergence$DaysAfterPlanting)
PlantEmergence$Rep <- as.factor(PlantEmergence$Rep)

```


## Question 2. 
Fit a linear model to predict Emergence using Treatment and
DaysAfterPlanting along with the interaction. Provide the summary of the linear
model and ANOVA results.

```{r}
Question2 <- lm(Emergence ~ Treatment * DaysAfterPlanting, data = PlantEmergence)

summary(Question2)

anova(Question2)
```


## Question 3. 
Based on the results of the linear model in question 2, do you need to fit the
interaction term? Provide a simplified linear model without the interaction term but still testing both main effects. Provide the summary and ANOVA results. Then, interpret the intercept and the coefficient for Treatment 2.

```{r}
Question2_simple <- lm(Emergence ~ Treatment + DaysAfterPlanting, data = PlantEmergence)

summary(Question2_simple)
anova(Question2_simple)
```

Results:
-   The intercept (182.163) represents the expected emergence for Treatment 1 at day 0 of DaysAfterPlanting  
-   The coefficient for Treatment 2 (-134.531) indicates that emergence is, on average 134.5 units lower then treatment 1 with DaysAfterPlanting being constant. 

## Question 4. 
Calculate the least square means for Treatment using the emmeans package and
perform a Tukey separation with the compact letter display using the cld function.
Interpret the results.

```{r}
emm <- emmeans(Question2_simple, ~ Treatment)
emm

cld_results <- cld(emm, adjust = "tukey")
cld_results
```

*Results:* 
-   Treatment 2 has significantly lower emergence than all other treatment groups. 
-   Treatment 7 has high emergence then treatment 2, but is lower than the top-performing treatment and significantly different from both the lowest and the highest group of emergence. 
-   Treatments 1, 4, 8, and 9 are not significant different from either treatment 7 or the highest emergence group. 
-   Treatments 3, 5, and 6 have the highest emergence but are not significantly different from treatments 1, 4, 8, and 9. 

## Question 5. 
The provided function lets you dynamically add a linear model plus one factor from
that model and plots a bar chart with letters denoting treatment differences. Use this model to generate the plot shown below. Explain the significance of the letters.

```{r}
plot_cldbars_onefactor <- function(lm_model, factor) {
  data <- lm_model$model
  variables <- colnames(lm_model$model)
  dependent_var <- variables[1]
  
  lsmeans <- emmeans(lm_model, as.formula(paste("~", factor))) # estimate lsmeans
  Results_lsmeans <- cld(lsmeans, alpha = 0.05, reversed = TRUE, details = TRUE, Letters = letters)
  
  # Extract letters
  sig.diff.letters <- data.frame(
    Results_lsmeans$emmeans[,1],
    str_trim(Results_lsmeans$emmeans[,7])
  )
  colnames(sig.diff.letters) <- c(factor, "Letters")
  
  # Prepare data for plotting
  ave_stand2 <- lm_model$model %>%
    group_by(!!sym(factor)) %>%
    dplyr::summarize(
      ave.emerge = mean(.data[[dependent_var]], na.rm = TRUE),
      se = sd(.data[[dependent_var]]) / sqrt(n())
    ) %>%
    left_join(sig.diff.letters, by = factor) %>%
    mutate(letter_position = ave.emerge + 10 * se)
  
  # Plot
  plot <- ggplot(data, aes(x = !!sym(factor), y = !!sym(dependent_var))) +
    stat_summary(fun = mean, geom = "bar") +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5) +
    ylab("Number of emerged plants") +
    geom_jitter(width = 0.02, alpha = 0.5) +
    geom_text(data = ave_stand2, aes(label = Letters, y = letter_position), size = 5) +
    xlab(as.character(factor)) +
    theme_classic()
  
  return(plot)
}
plot_cldbars_onefactor(Question2_simple, "Treatment")
```

Results: 
-   Letter 'a' includes treatments 3, 5, and 6 are have the highest emergence but are not significantly different between group 'ab' 
-   Letters 'ab' mean treatment 1, 4, 8 and 9 are not significantly different from each group a and group b as they share similar components of both groups. 
-   Letter 'b' includes treatment 7 that has the a intermediate level of emergence and is significant lower then group a and higher then group c. 
-   Letter 'c' includes treatment 2 that has the significantly  lowest level of emergence compared to groups a and b. 

## Question 6. 
Generate the gfm .md file along with a .html, .docx, or .pdf. Commit, and push the
.md file to github and turn in the .html, .docx, or .pdf to Canvas. Provide me a link here to your github.

[Link to CodingChallenge7 folder on GitHub]()



