---
title: "Coding Challenge 5"
author: "Madelynn VanderHoek"
date: "`r Sys.Date()`"
output: 
    html_document:
       toc: TRUE
    md_document: 
---

### Question 1.

Download two .csv files from Canvas called DiversityData.csv and Metadata.csv, and read them into R using relative file paths.

```{r}
DiversityData <- read.csv("DiversityData.csv")
head(DiversityData)

Metadata <- read.csv("Metadata.csv")
head(Metadata)

library(tidyverse)

```

### Question 2.

Join the two data frames together by the common column ‘Code’. Name the resulting data frame alpha.

```{r}
alpha <- left_join(Metadata, DiversityData, by = "Code")
head(alpha)

```

### Question 3.

Calculate Pielou’s evenness index: Pielou’s evenness is an ecological parameter calculated by the Shannon diversity index (column Shannon) divided by the log of the richness column.

```         
a.Using mutate, create a new column to calculate Pielou’s evenness index.

b. Name the resulting dataframe alpha_even.
```

```{r}
alpha_even <- mutate(alpha, PielouEvenness = shannon/log(richness))
head(alpha_even)

```

### Question 4.

Using tidyverse language of functions and the pipe, use the summaries function and tell me the mean and standard error evenness grouped by crop over time.

```         
a. Start with the alpha_even dataframe

b. Group the data: group the data by Crop and Time_Point.

c. Summarize the data: Calculate the mean, count, standard deviation, and standard error
for the even variable within each group.

d. Name the resulting dataframe alpha_average
```

```{r}
alpha_average <- alpha_even %>%
  group_by(Crop, Time_Point) %>%
  summarise(
    mean_even = mean(PielouEvenness), 
    n = n(), 
    sd.dev = sd(PielouEvenness)) %>%
  mutate(std.err = sd.dev/sqrt(n))

head(alpha_average)
```

### Question 5.

Calculate the difference between the soybean column, the soil column, and the difference between the cotton column and the soil column

```         
a. Start with the alpha_average dataframe
b. Select relevant columns: select the columns Time_Point, Crop, and mean.even.
c. Reshape the data: Use the pivot_wider function to transform the data from long to wide
format, creating new columns for each Crop with values from mean.even.
d. Calculate differences: Create new columns named diff.cotton.even and diff.soybean.even
by calculating the difference between Soil and Cotton, and Soil and Soybean, respectively.
e. Name the resulting dataframe alpha_average2
```

```{r}
alpha_average2 <- alpha_average %>%
  select(Time_Point, Crop, mean_even) %>%
  pivot_wider(names_from = Crop, values_from = mean_even) %>%
  mutate(diff.cotton.even = Soil - Cotton) %>%
  mutate(diff.soybean.even = Soil - Soybean)

head(alpha_average2)

```

### Question 6.

Connecting it to plots

```         
a. Start with the alpha_average2 dataframe
b. Select relevant columns: select the columns Time_Point, diff.cotton.even, and diff.soybean.even.
c. Reshape the data: Use the pivot_longer function to transform the data from wide to long
format, creating a new column named diff that contains the values from diff.cotton.even
and diff.soybean.even.

i. This might be challenging, so I’ll give you a break. The code is below. 
pivot_longer(c(diff.cotton.even, diff.soybean.even), names_to = "diff")

d. Create the plot: Use ggplot and geom_line() with ‘Time_Point’ on the x-axis, the column
‘values’ on the y-axis, and different colors for each ‘diff’ category. The column named
‘values’ come from the pivot_longer. The resulting plot should look like the one to the right.
```

```{r}
alpha_average_plot <- alpha_average2 %>%
  select(Time_Point, diff.cotton.even, diff.soybean.even) %>%
  pivot_longer(c(diff.cotton.even, diff.soybean.even), names_to = "diff") %>%
  ggplot(aes(x = Time_Point, y = value, color = diff)) +
  geom_line() +
  theme_minimal() +
  xlab("Time Point") +
  ylab("Values")

alpha_average_plot


```

### Question 7.

Commit and push a gfm .md file to GitHub inside a directory called Coding Challenge 5. Provide me a link to your github written as a clickable link in your .pdf or.docx.

[GitHub Link]<https://github.com/MadelynnV/PLAP-Clas-2026>
